---
output: 
  github_document:
    pandoc_args: --webtex
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

devtools::load_all()
```

#Introduction

Rfuels is a small package for estimating surface fuel loads from Brown's 
transects (a.k.a. line-intercept transects) in forests dominated by Sierra 
Nevada conifer species. 

The common protocols for sampling surface fuels in 
forest ecosystems directly describe the fuel load in terms of duff/litter 
depth, counts of fine woody fuels, and surveys of coarse woody debris. However, 
many uses of these data (e.g. fire behavior models and carbon accounting) 
require the fuel load to be expressed in terms of mass of surface fuels 
per unit area. 

Brown (1974) described a general-purpose set of equations to 
estimate fuel loads from transect data. Rfuels uses the species composition of 
the local overstory to improve the fuel load estimates from Brown's transects. 

See "Using Rfuels" for information about data requirements and using the 
package, and see "Background Information" for a more detailed description 
of the estimation method.

#Installing Rfuels

Someday, Rfuels may appear on the CRAN repository. For now, you can install 
Rfuels from the GitHub repo:

```{r eval = FALSE}
# install the 'devtools' package
install.packages("devtools")

# load the 'devtools' package
library(devtools)

# install Rfuels
devtools::install_github('danfosterfire/Rfuels')

```

Then you can load the Rfuels package with:

```{r}
library(Rfuels)
```

#Using Rfuels

Rfuels provides a high-level wrapper function, **estimate_fuel_loads()**, which  
takes three arguments:

  - *fuels_data*, a properly-formatted dataframe containing 
  the observed fuels data
  - *trees_data*, a properly-formatted containg the observed tree data
  - *results_type*, either 'full', 'results_only', or 'fuels_only'. This sets 
  the verbosity of the results dataframe.
  
**estimate_fuel_loads()** will return a tidy dataframe, with each row a unique 
observation (a fuels transect sampled on a specific date) and columns describing 
the surface fuel load in various subcategories. 

##Data Requirements

Rfuels imposes very specific requirements on the type and format of input data, 
so read this section carefully!

###Fuels Data

Surface fuels should be recorded in the field following Brown (1974) or a 
similar method. The specifics of transect lengths, number of transects 
per plot, and the number of litter and duff samples per transect are variable, 
but the broad outline is common:

  - Each sampling location (plot) will have one or more transects starting 
  at the plot center
  
  - All measurements are collected along all transects
  
  - Litter and duff depths are sampled at fixed location(s) along each transect
  
  - For predefined distances along the transect, the number of fine woody fuel 
  particles that intersect the transect is recorded as a count. Fine woody 
  fuel particles are categorized by size into 1-hour, 10-hour, and 100-hour 
  fuels, and recorded separately.
  
  - Each 1000-hour fuel intersecting the transect has its diameter and 
  decay class recorded individually
  
  - Most versions of the protocol also call for measuring the depth of the 
  entire fuelbed, but this information is not used for estimating the fuel 
  load represented by the sample
  
Rfuels requires the input fuels data to be formatted as a dataframe with a 
row for each observation (a transect on a specific date). The dataframe must 
have at least these columns (column names are exact):

  - **plot_id**: Plot IDs identify the sampling location. If you have 
  a nested study design with "stand A plot 1" and "stand B plot 1", use 
  "A-1" and "B-1" as PlotIDs. (This will be helpful when running statistical 
  analyses later.) There may be multiple inventory dates per
  plot_id and/or multiple transects sharing a plot_id. plot_id will be coerced 
  to a character vector on import.
  
  - **inv_date**: The date of measurement, in mm/dd/YYYY format. There must 
  be a 1:1 match between the dates of fuels measurements and dates of trees 
  data. 

  - **azimuth**: Integer in the range 0-359 (inclusive) giving the
 azimuth from plot center for the transect. Rfuels uses the azimuth as an 
 identifier for individual transects within plots. Must be unique within a
 (plot_id:inv_date). Will be coerced to character.

 - **x1h_length_m**, **x10h_length_m**, **x100h_length_m**, and 
 **x1000h_length_m**: Numeric(s) greater than 0. 
 The length (in meters) of the sampling transects for 1-hour,
 10-hour, 100-hour, and 1000-hour fuels, respectively. Transect lengths 
 may vary by plot_id and/or year, for example if sampling protocols changed 
 over time, or 1000-hour transects were extended until at least one 
 intersection was found.

  - **count_x1h, count_x10h, count_x100h**: Integers greater than or
 equal to 0. Transect counts of the number of intersections for 1-, 10-,
 and 100-hour fuels, respectively.

 - **duff_depth_cm** and **litter_depth_cm**: Numeric greater than 0.
 Average depth of duff / litter (in cm) on the transect. Users with multiple
 measurements per transect of litter and/or duff should average them
 together before import. Depths must be recorded in cm.

 - **sum_d2_1000r_cm2** and **sum_d2_1000s_cm2** The sum-of-squared-diameters
 for 1000-hour fuels on the transect, for rotten and sound fuels
 respectively. Users must aggregate their large fuels (1000-hour) into
 sound or rotten classes, and sum the squared diameters (in cm) for all
 1000-s or 1000-r intersections on the transect.

Additionally, the dataframe *may* have a column for **slope_percent**, the
slope (in percent) along the transect. Brown's equations include the
option to correct for the slope effect on horizontal length of transects.
*Keep in mind that this correction factor applies to the transect slope,
not the plot slope.* If a slope_percent is not supplied, we set the slope
correction factor to 1 (no slope).

###Trees Data

The dataframe must have a row for each observation of each individual
tree on each sampling date (this format is commonly called a "treelist"). The 
treelist must have at least these columns:

  - **plot_id**: Plot IDs identify the sampling location. If you have 
  a nested study design with "stand A plot 1" and "stand B plot 1", use full 
  nested identifiers (e.g. "A-1" and "B-1"). This will be helpful when running
  statistical analyses later.) There may be multiple inventory dates per
  plot_id and/or multiple transects sharing a plot_id. plot_id will be coerced 
  to a character vector on import.
  
  - **inv_date**: The date of measurement, in mm/dd/YYYY format. There must 
  be a 1:1 match between the dates of fuels measurements and dates of trees 
  data. 
  
  - **species**: A species identifier code for the individual tree. Generally 
  follows 4-letter scientific abbreviation format (e.g. Abies concolor is 
  "ABCO", not "WF", "White fir", "Abies_concolor", etc.). Compare your species 
  codes to those included in the Van Wagtendonk constant tables (see 
  "Background Information") to ensure correct matching. Note that 
  singleleaf pinyon (Pinus monophylla) and western white pine (Pinus monticola) 
  would share the code "PIMO" - these should be labeled "PIMO1" and "PIMO2",
  respectively.

  - **dbh_cm**: The diameter at breast height (4.5', 1.37m) of the tree in
  centimeters.

## Load Data

Our nice example data are .csv files which are already properly formatted:

```{r eval = FALSE}
# load the example fuels data from the .csv file
example_fuels_data = read.csv(file = 'path/to/your/data/fuels.csv',
                              stringsAsFactors = TRUE)

# load the example trees data from the .csv file
example_trees_data = read.csv(file = 'path/to/your/data/trees.csv',
                              stringsAsFactors = TRUE)
```

And here's what the example data look like:

```{r include = FALSE}
# load the example fuels data from the .csv file
example_fuels_data = 
  utils::read.csv(file = system.file('extdata', 'example_fuels.csv', 
                                      package = 'Rfuels'),
                  stringsAsFactors = TRUE)



# load the example trees data from the .csv file
example_trees_data = 
  utils::read.csv(file = system.file('extdata', 'example_treelist.csv',
                                     package = 'Rfuels'),
                  stringsAsFactors = TRUE)

```

```{r}
head(example_fuels_data)

head(example_trees_data)
```

##Call estimate_fuel_loads()

Once your input data are properly formatted, you can call estimate_fuel_loads() 
and assign the resulting dataframe to a variable:

```{r}
transect_fuel_loads =  
  
  estimate_fuel_loads(fuels_data = example_fuels_data,
                      trees_data = example_trees_data,
                      results_type = 'results_only')


```

We got a message about the type of results returned, and also a warning 
about our input treelist. Our dataset includes *Quercus kelloggii*, which isn't 
one of the tree species for which we have empirical data to plug into the fuel 
load estimations. **estimate_fuel_loads()** makes a best-guess and counts all 
unrecognized tree species as 'OTHER', and applies the "All species" generic 
coefficients from Van Wagtendonk et al. (1996) and (1998). 

(How appropriate it is to apply an 'all species' constant for Sierra Nevada 
conifers to black oak fuels is another question; hopefully in the future 
we can expand the empirical dataset Rfuels draws upon.)

Now we have a dataframe with the fuel load (and some overstory information) 
for each transect in each year:

```{r}
head(transect_fuel_loads)
```

##Results details

The results_only dataframe (the default) includes the following columns:

  - **plot_id**: A factor with the plot_id (location in space and/or study 
  design) for each observation.
  
  - **inv_date**: The date of the observation.
  
  - **pBA_[species]**: For each [species], the proportion of total plot basal 
  area occupied by that species. For example, if a plot has many small 
  white fir totalling 0.06 $m^2$ basal area, and a few large ponderosa pine 
  totalling 0.14 $m^2$ basal area, *pBA_ABCO* = 0.3 and *pBA_PIPO* = 0.7. These 
  columns describe the species composition of the overstory for the transect.
  
  - **pBA_(all)**: The total proportional basal area of all species on the plot, 
  which should always be equal to 1. (Or nearly so, within a small rounding 
  error.)
  
  - **azimuth**: The azimuth from plot center used for the transect. This can 
  (and should) be used as a nested observation identifier within a plot_id.
  
  - **fuelload_litter_Mgha**: The fuel load of litter, in metric tons per 
  hectare, as sampled on a single transect. Robust stand-level estimates 
  should include many transects.
  
  - **fuelload_duff_Mgha**: The fuel load of duff, in metric tons per hectare.
  
  - **fuelload_1h_Mgha**: The fuel load of 1-hour fuels (0 - 0.64 cm diameter), 
  in metric tons per hectare.
  
  - **fuelload_10h_Mgha**: The fuel load of 10-hour fuels (0.64 - 2.54 cm 
  diameter), in metric tons per hectare.
  
  - **fuelload_100h_Mgha**: The fuel load of 100-hour fuels (2.54 - 7.62 cm 
  diameter), in metric tons per hectare.
  
  - **fuelload_1000s_Mgha**: The fuel load of sound 1000-hour fuels (7.62+ cm 
  diameter) in metric tons per hectare.
  
  - **fuelload_1000r_Mgha**: The fuel load of rotten 1000-hour fuels (7.62+ cm 
  diameter) in metric tons per hectare.
  
  - **fuelload_fwd_Mgha**: The total 'fine woody debris' fuel load (1-hour, 
  10-hour, and 100-hour fuels combined) in metric tons per hectare.
  
  - **fuelload_1000h_Mgha**: The total 1000-hour fuel load (sound and rotten 
  combined) in metric tons per hectare.
  
  - **fuelload_surface_Mgha**: The total surface fuel load, in metric tons 
  per hectare. This is the sum of the litter, fine woody debris, and 1000-hour 
  fuel loads.
  
  - **fuelload_total_Mgha**: The total fuel load, in metric tons per hectare. 
  This is the sum of the litter, duff, fine woody debris, and 1000-hour fuel 
  loads.

The *results_full* dataframe includes all of the above, plus these additonal 
columns describing the directly observed values and some intermediate 
calculations:

  - **x1h_length_m**, **x10h_length_m**, **x100h_length_m**, and 
  **x_1000h_length_m**: Relevant transect lengths in meters.
  
  - **count_x1h**, **count_x10h**, **count_x100h**, **duff_depth_cm**, 
  **litter_depth_cm**, **sum_d2_1000r_cm2**, and **sum_d2_1000s_cm2**: The 
  empirical observations for each fuel category, as supplied by the input data.
  
  - **slp_c**: The slope correction factor used for the transect. Defaults to 
  1 if no transect-level slope information is available. See 
  "Background Information" for more details.
  
  - **litter_coeff** and **duff_coeff**: Plot-specific estimates of the 
  linear coefficient between duff / litter depth (in cm) and duff / litter 
  fuel load (in kg / $m^2$). Derived by averaging species-specific 
  empirical values according to the species composition of the plot overstory. 
  See "Background Information" for more details.
  When multiple transects share the same plot_id, they will have the same 
  overstory, and will share these coefficients.
  
  - **x1h_coeff**, **x10h_coeff**, and **x100h_coeff**: Plot-specific 
  coefficients derived by averaging species-specific 
  empirical values according to the species composition of the plot 
  overstory. See "Background Information" for more details.
  When multiple transects share the same plot_id, they will have the same 
  overstory, and will share these coefficients.
  
  - **x1000s_coeff** and **x1000r_coeff**: A plot-specific 
  coefficients derived by averaging species-specific 
  empirical values according to the species composition of the plot 
  overstory. See "Background Information" for more details.
  When multiple transects share the same plot_id, they will have the same 
  overstory, and will share these coefficients.
  
Finally, the *fuels_only* dataframe includes only the observation identification
columns (*plot_id*, *inv_date*, and *azimuth*) and the columns for the 
various fuel load categories.

##Working with fuel load data

*under construction*

#Background Information

This method for calculating fuel loads from Brown's transects was originally 
described in Stephens (2001). Specifically:

> "Surface and ground fuel loads were calculated by using appropriate equations 
developed for Sierra Nevada forests (van Wagtendonk et al. 1996, 1998). 
Coefficients required to calculate all surface and ground fuel loads were 
arithmetically weighted by the basal area fraction (percentage of total basal 
area by species) to produce accurate estimates of fuel loads (Jan van 
Wagtendonk, personal communication, 1999)." 

Since then, the method has been used repeatedly in other publications, 
particularly those related to the Fire-Fire-Surrogate study at Blodgett Forest 
Reserach Station, e.g. Stephens and Moghaddas (2005) and Stephens 
et al. (2012). At some point, Jason Moghaddas constructed an excel workbook 
(with included macros) to facilitate use of the method for new datasets. 

The this package implements the method in R, using Moghaddas' 
implementation (and various publications) for reference. The goal of this 
effort is to make the method more accessible, transparent, and reproducible 
for use in future research.

This method requires Brown's transect data and plot (or stand) level overstory 
data. Brown (1974) describes the widely used fuels transect sampling protocol 
and provides equations used to calculate fuel loads for woody debris from the 
transect samples. 

The general idea of Stephens's modification of the method is to refine the 
fuel load estimates by using overstory data to improve the accuracy of 
coefficents used in Brown's equations to calculate 
fuel loads from transect data. Specifically, QMD, SEC, and SG are coefficients 
which vary by the species that is the fuel source, and so better estimates can 
(presumably) be derived by using species-weighted-average values for QMD, SEC, 
and SG rather than using Brown's given values, which are generalized for 
entire regions. (See below for definitions of QMD, SEC, and SG.)

**Note:** This implementation of the method assumes the user is working in the 
Sierra Nevada. Any species not included in the Van Wagtendonk et al. tables 
below is assigned their "All Species" value by default. Also, per the usage of 
Moghaddas' spreadsheet for previous studies, I have elected to include both 
live and dead trees in the basal-area calculations for the purposes of 
calculating fuel loads.

This method breaks forest fuels into three main categories, each of which has a 
particular implementation for calculating the fuel load represented by transect 
samples.

##Litter and Duff

Litter and duff are measured as depths as specific points along a sampling 
transect. Van Wagtendonk (1998) developed regressions for litter, Duff, and 
combined-litter-and-duff loading (kg/$m^2$) as a function of depth (cm) for 19 
different Sierra Nevada conifer species: 

```{r echo = FALSE}
knitr::kable(Rfuels:::litterduff_coeffs,
             caption = 'Van Wagtendonk et al. (1998) Table 7: Regression statistics for litter, duff, and litter and duff weight (kg m-2) as a function of their respective depths (cm) for 19 Sierra Nevada conifers')
```

The fuel load represented by a depth measurement under a mixed-species overstory can be estimated using the equation

$$F_{d,plot} = d * Coeff_{plot}$$

where:

  - $F_{d,plot}$ is the fuel load (in kg/$m^2$)

  - $d$ is the depth of litter, duff, or litter and duff together (in cm) at some 
  point along the transect. If depth was taken at multiple points along the 
  transect, average them together to calculate *d*.

  - $Coeff_{plot}$ is the best estimate coefficient for the linear relationship 
  between depth and fuel load for a fuel bed generated by the overstory present 
  at *plot*

We can calculate $Coeff_{plot}$ by averaging together the different species-specific 
coefficients for each tree species contributing fuel to the plot, weighted by 
their local prevalence. Specifically, we weight each species' coefficient by 
the proportion of total basal area contributed by that species:

$$Coeff_{plot} = \sum_{spp}{[(\frac{BA_{spp}}{BA_{total}})*Coeff_{spp}]}$$

Where 

  - $Coeff_{spp}$ is the species-specific coefficient for species *spp*, 
  
  - $Coeff_{plot}$ is the best estimate multiple-species coefficient we will 
  use to estimate the fuel load for transects taken in the plot

  - $BA_{spp}$ and $BA_{total}$ are the basal area occupied by species *spp* 
  and the total overstory basal area (respectively) in the plot. 
  
  - $\frac{BA_{spp}}{BA_{total}}$ is proportion of the plot's basal area 
  occupied by species *spp*. 

This esimate of the fuel load represented by the transect can be averaged with 
other transects at the same plot (which will have the same value for 
$Coeff_{plot}$, but different values for *d*) to generate plot-level 
estimates of fuel load, which can themselves be averaged to generate unit-level 
estimates of fuel load, depending on the user's study design.

##Fine Woody Debris: 1-hour, 10-hour, and 100-hour fuels

Calculating fuel loads represented by transect counts of 1-hour, 10-hour, and 
100-hour fuels is more complicated, but follows the same principle as described 
for litter and duff above. The different timelag classes (which correspond to 
size classes of 0.0-0.64cm, 0.64-2.54cm, and 2.54-7.62cm, respectively) must 
be calculated separately but in a common manner.

Van Wagtendonk et al. (1996) give the equation (modified from Brown 1974):

$$W = \frac{const*QMD*SEC*SLP*SG*n}{length}$$

Where:

  - $W$ is the estimated weight (in tons/acre, kg/ha, etc.) of fine woody 
  debris size class *f* on the plot 

  - $const$ is a constant composed of various unit conversion factors (e.g., 
  to get to tons/acre in the end), described by some authers as *k*

  - $QMD$ is the quadratic mean diameter of the fuels recorded in the transect

  - $SEC$ is the correction for the nonhorizontal angle of the fuels

  - $SLP$ is the slope correction factor, which allows corrected for the effect 
  of a non-horizontal transect on the horizontal distance sampled

  - $SG$ is the specific gravity of the fuels in the transect

  - $n$ is the number of fuel particles intersecting the transect,

  - and $length$ is the length of the transect. 

Procedures for calculating these values are described below. *QMD*, *SEC*, and 
*SG* all vary by timelag classification, and so *W* is calculated 
individually for each timelag (1h, 10h, and 100h).

###Const (k-values)

Equation constant *k* ("const" in van Wagtendonk 1996 and Brown 1974) for 
sampled units and output units. These values for *k* are from van Wagner 
(1982), and these are used by forest service per Woodall and Williams (2008).
(These values are also consistent with Moghaddas' work.)

```{r echo = FALSE}
knitr::kable(Rfuels:::kvals,
             caption = 'Van Wagner (1982) Table 1: Equation constant k for some length, volume, and weight units in line intersect sampling')
```



###QMD (Quadratic mean diameter in centimeters squared)

QMD, SEC, and SG vary by species and timelag classification of the fuel being 
totaled. Values for specific species/timelag for fine woody fuels are drawn 
from van Wagtendonk (1996).

```{r echo = FALSE}
knitr::kable(Rfuels:::QMDcm,
             caption = 'Van Wagtendonk et al. (1996) Table 3: Average squared quadratic mean diameter by fuel size class for 19 Sierra Nevada conifers')
```

These constants are used in combination with overstory data to create an aggregate estimate of QMD, an average of the various species' QMD estimates (from Van Wagtendonk et al.) weighted by the proportion of stand basal area occupied by each species. See the following formula:

$$QMD_{plot,timelag} = \sum_{species=spp}{PropBA_{spp,plot}*QMD_{spp,timelag}}$$

Where 

  - $QMD_{plot}$ is the estimated quadratic mean diameter of fuels in the 
  *timelag* class in the *plot* across all species

  - $PropBA_{spp,plot}$ is the proportion of total basal area in *plot* occupied 
  by the species *spp*, calculated from the measurement data

  - $QMD_{spp,timelag}$ is the QMD of fuels in the *timelag* class for species 
  *spp*, as reported by Van Wagtendonk et al. (1996) Table 3.

###SEC (secant of acute angle)

```{r echo = FALSE}
knitr::kable(Rfuels:::SEC,
             caption = 'Van Wagtendonk et al. (1996) Table 6: Average secant of acute angles of inclination of nonhorizontal particles by fuel size class for 19 Sierra Nevada conifers')
```

A propotion-BA-weighted average of SEC is generated in the same way as QMD above.

###SG (specific gravity)

Note that van Wagtendonk et al. found that "species was not significant for the 
7.62+cm (3+ in) rotten fuels." and that the average specific gravity for rotten 
fuels was .36, which is here included as a value.

```{r echo = FALSE}
knitr::kable(Rfuels:::SG,
             caption = 'Van Wagtendonk et al. (1996) Table 8: Average specific gravity by fuel size class for 19 Sierra Nevada conifers')
```

A propotion-BA-weighted average of SEC is generated in the same way as QMD above.

###SLP (slope correction factor)

*SLP* varies by the plot location and transect 
azimuth, $SLP = c = \sqrt{1+(\frac{percentslope}{100})^2}$ per Brown 1974. 
This is a simple adjustment for the influence of slope on transect length.

###length 

*length* is the length of the sampling transect, and varies by the sampling 
protocol (which may vary from year-to-year) and the timelag class.

###n 

*n* is the number of intersections of the given timeclass in the plane of the 
transect.

##Coarse Woody Debris: 1000-hour fuels

Brown (1974) gives:

$$W_{1000h} = \frac{11.64*\sum{d^2}*s*a*c}{N*l}$$

and notes: 

> "For material 3 inches and larger, square the diameter of each intersected 
piece and sum the squared values ($\sum{d^2}$) for all pieces in the sampled 
area. Compute $\sum{d^2}$ separately for sound and rotten categories. To 
obtain weights or volumes for certain diameter ranges (3 to 9 inches, for 
example), compute $\sum{d^2}$ for the specified range."

This equation is just a special case of the equations given above for 1-100 
hour fuels. 11.64 is a unit conversion factor for US units (*const* as above), 
*s* is the specific gravity of the fuel (*SG* as above), 
*a* is a value for the secant angle (*SEC* as above), and *c* is the slope 
correction factor (*SLP* as above). *N* is the number of transects 
represented in the calculation, and is assumed to be 1 in this method. 
*l* is the transect length, as described above.

The difference is that instead of counted intercepts and an average squared 
quadratic mean diameter, we have the actual sum of squared diameters from the 
measurements directly. 

For sound 1000-hour fuels, we can substitute the BA-weighted-average for a 
specific transect's overstory and re-arrange the equation:

$$W_{1000S} = (\sum{d^2})*(\frac{const*SLP}{length})*\sum_{spp}{(\frac{BA_{spp}}{BA_{total}})*SEC_{spp,1000s})}*(\sum_{spp}{(\frac{BA_{spp}}{BA_{total}})*SG_{spp,1000s}})$$

van Wagtendonk 1998 give species-specific values for QMD, SEC, and SG, for both 
1000-hour sound and 1000-hour rotten fuels. 

Users who have diameter measurements for 1000-hour fuels (which is a 
standard sampling protocol) have actual diameter measurements, and don't need 
the by-species QMD averages from Van Wagtendonk et al. (1996). 

Van Wagtendonk et al. (1996) do not discuss sound vs. rotten 1000-hour fuels, 
stating only that "The average secant of the acute angle to the horizontal for 
the 7.62+ cm (3+in) size class for all species was 2.67 (Table 6)." However, 
actually referring to table six gives very different data, with by-species 
averages recorded and varying from 1.00 to 1.06, with an average of 1.02. I 
believe the text is in error (2.67 bears no resemblance to the data given in 
the table, but it is 7.62 reversed.) 

For SG, "The average specific gravity for rotten fuels was .36." and "species 
was not significant for the 7.62+ cm (3+ in) rotten fuels." Specific gravities 
by species for sound 1000-hour fuels are given in Van Wagtendonk et al. (1996), 
Table 8. 

#References

  - Brown, J. K. *Handbook for Inventorying Downed Woody Material.* USDA For. 
  Serv. Gen. Tech. Rep. 24 (1974). https://www.fs.usda.gov/treesearch/pubs/28647

  - Stephens, S. L. *Fire history differences in adjacent Jeffrey pine and upper 
  montane forests in the eastern Sierra Nevada.* Int. J. Wildl. Fire 10, 161–167 
  (2001). https://doi.org/10.1071/WF01008
  
  - Stephens, S. L. & Moghaddas, J. J. *Experimental fuel treatment impacts on 
  forest structure, potential fire behavior, and predicted tree mortality in a 
  California mixed conifer forest.* For. Ecol. Manage. 215, 21–36 (2005). 
  https://doi.org/10.1016/j.foreco.2005.03.070
  
  - Stephens, S. L., Collins, B. M. & Roller, G. *Fuel treatment longevity in 
  a Sierra Nevada mixed conifer forest.* For. Ecol. Manage. 285, 204–212 (2012).
  https://doi.org/10.1016/j.foreco.2012.08.030
  
  - Van Wagner, C. E. *Practical aspects of the line intersect method.* Forestry 
  18 (1982). http://www.cfs.nrcan.gc.ca/pubwarehouse/pdfs/6862.pdf
  
  - Van Wagtendonk, J. W., Benedict, J. M. & Sydoriak, W. M. *Physical 
  properties of woody fuel particles of Sierra Nevada conifers.* Int. J. Wildl. 
  Fire 6, 117–123 (1996). https://doi.org/10.1071/WF9960117
  
  - Van Wagtendonk, J. W., Benedict, J. M. & Sydoriak, W. M. *Fuel bed 
  characteristics of Sierra Nevada conifers.* West. J. Appl. For. 13, 73–84 
  (1998). https://doi.org/10.1093/wjaf/13.3.73
  

  
